Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms

### 2023-06-02 output/latex/manuscript.tex
latex.yaml: add variables
update manuscript.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{mathtools}
\usepackage{lmodern}
\usepackage{iftex}
\usepackage[legalpaper, left=0.5in, right=0.5in, top=1in, bottom=1in]{geometry}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\usepackage[inkscapeformat=pdf,inkscapelatex=false]{svg}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{american}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}

@@

Genes work together in context-specific networks to carry out different functions {[}\protect\hyperlink{ref-e0tRKjE5}{1},\protect\hyperlink{ref-CVF61Un5}{2}{]}.
Variations in these genes can change their functional role and, at a higher level, affect disease-relevant biological processes {[}\protect\hyperlink{ref-Om8ZhS06}{3}{]}.
In this context, determining how genes influence complex traits requires mechanistically understanding expression regulation across different cell types {[}\protect\hyperlink{ref-9Pr9idng}{4},\protect\hyperlink{ref-1DoyZS7y0}{5},\protect\hyperlink{ref-nhaocxmR}{6}{]}, which in turn should lead to improved treatments {[}\protect\hyperlink{ref-REXpV7nA}{7},\protect\hyperlink{ref-PgEwSS4Q}{8}{]}.
Previous studies have described different regulatory DNA elements {[}\protect\hyperlink{ref-1DoyZS7y0}{5},\protect\hyperlink{ref-15J98V2qM}{9},\protect\hyperlink{ref-sLkFMFZj}{10},\protect\hyperlink{ref-SxuuTQTQ}{11},\protect\hyperlink{ref-xRGqPsT2}{12}{]} including genetic effects on gene expression across different tissues {[}\protect\hyperlink{ref-9Pr9idng}{4}{]}.
Integrating functional genomics data and GWAS data {[}\protect\hyperlink{ref-WFslDIWl}{13},\protect\hyperlink{ref-WFslDIWl}{13},\protect\hyperlink{ref-aIyQY5ZT}{14},\protect\hyperlink{ref-z8MQTAnJ}{15}{]} has improved the identification of these transcriptional mechanisms that, when dysregulated, commonly result in tissue- and cell lineage-specific pathology {[}\protect\hyperlink{ref-O0e3EhY6}{16},\protect\hyperlink{ref-18jYvPauB}{17},\protect\hyperlink{ref-wNE0EQlN}{18}{]}.

Given the availability of gene expression data across several tissues {[}\protect\hyperlink{ref-9Pr9idng}{4},\protect\hyperlink{ref-6SPTvFXq}{19},\protect\hyperlink{ref-Nz3IMEzd}{20},\protect\hyperlink{ref-57TjOMEA}{21}{]}, an effective approach to identify these biological processes is the transcription-wide association study (TWAS), which integrates expression quantitative trait loci (eQTLs) data to provide a mechanistic interpretation for GWAS findings.
TWAS relies on testing whether perturbations in gene regulatory mechanisms mediate the association between genetic variants and human diseases {[}\protect\hyperlink{ref-yEdRP9Xx}{22},\protect\hyperlink{ref-1D63fEEPb}{23},\protect\hyperlink{ref-1FFzCXo1s}{24},\protect\hyperlink{ref-Z8bvDdVq}{25}{]}, and these approaches have been highly successful not only in understanding disease etiology at the transcriptome level {[}\protect\hyperlink{ref-ktVcsmYD}{26},\protect\hyperlink{ref-gZAOkumx}{27},\protect\hyperlink{ref-zwpq2IXD}{28}{]} but also in disease-risk prediction (polygenic scores) {[}\protect\hyperlink{ref-hSYqogYZ}{29}{]} and drug repurposing {[}\protect\hyperlink{ref-17oeJ0CXy}{30}{]} tasks.
However, TWAS works at the individual gene level, which does not capture more complex interactions at the network level.

These gene-gene interactions play a crucial role in current theories of the architecture of complex traits, such as the omnigenic model {[}\protect\hyperlink{ref-vpIDZCSa}{31}{]}, which suggests that methods need to incorporate this complexity to disentangle disease-relevant mechanisms.
Widespread gene pleiotropy, for instance, reveals the highly interconnected nature of transcriptional networks {[}\protect\hyperlink{ref-pZZn28he}{32},\protect\hyperlink{ref-PDWEwciL}{33}{]}, where potentially all genes expressed in disease-relevant cell types have a non-zero effect on the trait {[}\protect\hyperlink{ref-vpIDZCSa}{31},\protect\hyperlink{ref-LXvTZzEA}{34}{]}.
One way to learn these gene-gene interactions is using the concept of gene module: a group of genes with similar expression profiles across different conditions {[}\protect\hyperlink{ref-CVF61Un5}{2},\protect\hyperlink{ref-jrAMOJCD}{35},\protect\hyperlink{ref-SjGoBywE}{36}{]}.
In this context, several unsupervised approaches have been proposed to infer these gene-gene connections by extracting gene modules from co-expression patterns {[}\protect\hyperlink{ref-S4e4WaP3}{37},\protect\hyperlink{ref-f2r8LLIn}{38},\protect\hyperlink{ref-VzZoy0BD}{39}{]}.
Matrix factorization techniques like independent or principal component analysis (ICA/PCA) have shown superior performance in this task {[}\protect\hyperlink{ref-1BVbSrr6M}{40}{]} since they capture local expression effects from a subset of samples and can handle modules overlap effectively.
Therefore, integrating genetic studies with gene modules extracted using unsupervised learning could further improve our understanding of disease origin {[}\protect\hyperlink{ref-SjGoBywE}{36}{]} and progression {[}\protect\hyperlink{ref-S56q1qoc}{41}{]}.

Here we propose PhenoPLIER, an omnigenic approach that provides a gene module perspective to genetic studies.
The flexibility of our method allows integrating different data modalities into the same representation for a joint analysis.
In this work, we show that this module perspective can infer how groups of functionally-related genes influence complex traits, detect shared and distinct transcriptomic properties among traits, and predict how pharmacological perturbations affect genes' activity to exert their effects.
PhenoPLIER maps gene-trait associations and drug-induced transcriptional responses into a common latent representation.
For this, we integrated thousands of gene-trait associations (using TWAS from PhenomeXcan {[}\protect\hyperlink{ref-lY5ln3dB}{42}{]}) and transcriptional profiles of drugs (from LINCS L1000 {[}\protect\hyperlink{ref-F7lIlh2N}{43}{]}) into a low-dimensional space learned from public gene expression data on tens of thousands of RNA-seq samples (recount2 {[}\protect\hyperlink{ref-6SPTvFXq}{19},\protect\hyperlink{ref-14rnBunuZ}{44}{]}).
We used a latent representation defined by a matrix factorization approach {[}\protect\hyperlink{ref-14rnBunuZ}{44},\protect\hyperlink{ref-Ki2ij7zE}{45}{]} that extracts gene modules with certain sparsity constraints and preferences for those that align with prior knowledge (pathways).
When mapping gene-trait associations to this reduced expression space, we observed that diseases were significantly associated with gene modules expressed in relevant cell types: such as hypothyroidism with T cells, corneal endothelial cells with keratometry measurements, hematological assays on specific blood cell types, plasma lipids with adipose tissue, and neuropsychiatric disorders with different brain cell types.
Moreover, since PhenoPLIER can use models derived from large and heterogeneous RNA-seq datasets, we could also identify modules associated with cell types under specific stimuli or disease states.
We observed that significant module-trait associations in PhenomeXcan (our discovery cohort) replicated in the Electronic Medical Records and Genomics (eMERGE) network phase III {[}\protect\hyperlink{ref-gZAOkumx}{27},\protect\hyperlink{ref-wfqjCerX}{46}{]} (our replication cohort).
Furthermore, we performed a CRISPR screen to analyze lipid regulation in HepG2 cells.
We observed more robust trait associations with modules than with individual genes, even when single genes known to be involved in lipid metabolism did not reach genome-wide significance.
Compared to a single-gene approach, our module-based method also better predicted FDA-approved drug-disease links by capturing tissue-specific pathophysiological mechanisms linked with the mechanism of action of drugs (e.g., niacin with cardiovascular traits via a known immune mechanism).
This improved drug-disease prediction suggested that modules may provide a better means to examine drug-disease relationships than individual genes.
Finally, exploring the phenotype-module space revealed stable trait clusters associated with relevant tissues, including a complex branch involving lipids with cardiovascular, autoimmune, and neuropsychiatric disorders.
In summary, instead of considering single genes associated with different complex traits, PhenoPLIER incorporates groups of genes that act together to carry out different functions in specific cell types.
This approach improves robustness in detecting and interpreting genetic associations, and here we show how it can prioritize alternative and potentially more promising candidate targets even when known single gene associations are not detected.
The approach represents a conceptual shift in the interpretation of genetic studies.
It has the potential to extract mechanistic insight from statistical associations to enhance the understanding of complex diseases and their therapeutic modalities.

@@

\begin{figure}
\hypertarget{fig:entire_process}{%
\centering
\includesvg[width=1\textwidth,height=\textheight]{images/entire_process/entire_process.svg}
\caption{\textbf{Schematic of the PhenoPLIER framework.}
\textbf{a)} High-level schematic of PhenoPLIER (a gene module-based method) in the context of TWAS (single-gene) and GWAS (single-variant).
In GWAS, we identify variants associated with traits.
In TWAS, first, we identify variants that are associated with gene expression levels (eQTLs); then, prediction models based on eQTLs are used to impute gene expression, which is used to compute gene-trait associations.
Resources such as LINCS L1000 provide information about how a drug perturbs gene expression; at the bottom-right corner, we show how a drug downregulates two genes (A and C).
In PhenoPLIER, these data types are integrated using groups of genes co-expressed across one or more conditions (such as cell types) that we call gene modules or latent variables/LVs.
\textbf{b)} The integration process in PhenoPLIER uses low-dimensional representations (matrices \(\mathbf{Z}\) and \(\mathbf{B}\)) learned from large gene expression datasets (top).
We used gene-drug information \(\mathbf{L}\) from LINCS L1000 and gene-trait associations \(\mathbf{M}\) from TWAS: PhenomeXcan was used as the discovery cohort, and eMERGE as replication (middle).
PhenoPLIER provides three computational components (bottom):
1) an LV-based regression model that associates an LV \(j\) (\(\mathbf{Z}_j\)) with a trait \(i\) (\(\mathbf{M}_i\)),
2) a clustering framework that learns groups of traits from TWAS associations projected into the LV space (\(\hat{\mathbf{M}}\)),
and 3) an LV-based drug repurposing approach that uses the projection of TWAS (\(\hat{\mathbf{M}}\)) and LINCS L1000 (\(\hat{\mathbf{L}}\)) into the LV space.
\textbf{c)} Genes that are part of LV603, termed as a neutrophil signature {[}\protect\hyperlink{ref-14rnBunuZ}{44}{]}, were expressed in relevant cell types (top).
LV603 was associated in PhenoPLIER with neutrophil counts and other white blood cells (bottom, showing the top 10 traits for LV603 after projecting gene-trait associations in PhenomeXcan).
eQTLs: expression quantitative trait loci;
MVN: multivariate normal distribution;
PBMC: peripheral blood mononuclear cells;
mDCs: myeloid dendritic cells.}\label{fig:entire_process}
}
\end{figure}

@@

We used TWAS results from PhenomeXcan {[}\protect\hyperlink{ref-lY5ln3dB}{42}{]} and the eMERGE network {[}\protect\hyperlink{ref-gZAOkumx}{27}{]} as discovery and replication cohorts, respectively (\protect\hyperlink{sec:methods:twas}{Methods}).
PhenomeXcan provides gene-trait associations for 4,091 different diseases and traits from the UK Biobank {[}\protect\hyperlink{ref-nmJxPpE5}{47}{]} and other studies, whereas the analyses on eMERGE were performed across 309 phecodes.
TWAS results were derived using two statistical methods (see \protect\hyperlink{sec:methods:predixcan}{Methods}):
1) Summary-MultiXcan (S-MultiXcan) associations were used for the regression and clustering components,
and 2) Summary-PrediXcan (S-PrediXcan) associations were used for the drug repurposing component.
In addition, we also used colocalization results, which provide a probability of overlap between the GWAS and eQTL signals.
For the drug-repurposing approach, we used transcriptional responses to small molecule perturbations from LINCS L1000 {[}\protect\hyperlink{ref-F7lIlh2N}{43}{]} comprising 1,170 compounds.

The latent gene expression representation was obtained from the MultiPLIER models {[}\protect\hyperlink{ref-14rnBunuZ}{44}{]}, which were derived by applying a matrix factorization method (the pathway-level information extractor or PLIER {[}\protect\hyperlink{ref-Ki2ij7zE}{45}{]}) to recount2 {[}\protect\hyperlink{ref-6SPTvFXq}{19}{]} -- a uniformly-curated collection of transcript-level gene expression quantified by RNA-seq in a large, diverse set of samples collected across a range of disease states, cell types differentiation stages, and various stimuli (see \protect\hyperlink{sec:methods:multiplier}{Methods}).
The MultiPLIER models extracted 987 LVs by optimizing data reconstruction but also the alignment of LVs with prior knowledge/pathways.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  legalpaperpaper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{mathtools}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=1in,left=0.5in,right=0.5in,bottom=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\usepackage[inkscapeformat=pdf,inkscapelatex=false]{svg}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{american}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}

@@

Genes work together in context-specific networks to carry out different functions\textsuperscript{\protect\hyperlink{ref-e0tRKjE5}{1},\protect\hyperlink{ref-CVF61Un5}{2}}.
Variations in these genes can change their functional role and, at a higher level, affect disease-relevant biological processes\textsuperscript{\protect\hyperlink{ref-Om8ZhS06}{3}}.
In this context, determining how genes influence complex traits requires mechanistically understanding expression regulation across different cell types\textsuperscript{\protect\hyperlink{ref-9Pr9idng}{4}--\protect\hyperlink{ref-nhaocxmR}{6}}, which in turn should lead to improved treatments\textsuperscript{\protect\hyperlink{ref-REXpV7nA}{7},\protect\hyperlink{ref-PgEwSS4Q}{8}}.
Previous studies have described different regulatory DNA elements\textsuperscript{\protect\hyperlink{ref-1DoyZS7y0}{5},\protect\hyperlink{ref-15J98V2qM}{9}--\protect\hyperlink{ref-xRGqPsT2}{12}} including genetic effects on gene expression across different tissues\textsuperscript{\protect\hyperlink{ref-9Pr9idng}{4}}.
Integrating functional genomics data and GWAS data\textsuperscript{\protect\hyperlink{ref-WFslDIWl}{13},\protect\hyperlink{ref-WFslDIWl}{13}--\protect\hyperlink{ref-z8MQTAnJ}{15}} has improved the identification of these transcriptional mechanisms that, when dysregulated, commonly result in tissue- and cell lineage-specific pathology\textsuperscript{\protect\hyperlink{ref-O0e3EhY6}{16}--\protect\hyperlink{ref-wNE0EQlN}{18}}.

Given the availability of gene expression data across several tissues\textsuperscript{\protect\hyperlink{ref-9Pr9idng}{4},\protect\hyperlink{ref-6SPTvFXq}{19}--\protect\hyperlink{ref-57TjOMEA}{21}}, an effective approach to identify these biological processes is the transcription-wide association study (TWAS), which integrates expression quantitative trait loci (eQTLs) data to provide a mechanistic interpretation for GWAS findings.
TWAS relies on testing whether perturbations in gene regulatory mechanisms mediate the association between genetic variants and human diseases\textsuperscript{\protect\hyperlink{ref-yEdRP9Xx}{22}--\protect\hyperlink{ref-Z8bvDdVq}{25}}, and these approaches have been highly successful not only in understanding disease etiology at the transcriptome level\textsuperscript{\protect\hyperlink{ref-ktVcsmYD}{26}--\protect\hyperlink{ref-zwpq2IXD}{28}} but also in disease-risk prediction (polygenic scores)\textsuperscript{\protect\hyperlink{ref-hSYqogYZ}{29}} and drug repurposing\textsuperscript{\protect\hyperlink{ref-17oeJ0CXy}{30}} tasks.
However, TWAS works at the individual gene level, which does not capture more complex interactions at the network level.

These gene-gene interactions play a crucial role in current theories of the architecture of complex traits, such as the omnigenic model\textsuperscript{\protect\hyperlink{ref-vpIDZCSa}{31}}, which suggests that methods need to incorporate this complexity to disentangle disease-relevant mechanisms.
Widespread gene pleiotropy, for instance, reveals the highly interconnected nature of transcriptional networks\textsuperscript{\protect\hyperlink{ref-pZZn28he}{32},\protect\hyperlink{ref-PDWEwciL}{33}}, where potentially all genes expressed in disease-relevant cell types have a non-zero effect on the trait\textsuperscript{\protect\hyperlink{ref-vpIDZCSa}{31},\protect\hyperlink{ref-LXvTZzEA}{34}}.
One way to learn these gene-gene interactions is using the concept of gene module: a group of genes with similar expression profiles across different conditions\textsuperscript{\protect\hyperlink{ref-CVF61Un5}{2},\protect\hyperlink{ref-jrAMOJCD}{35},\protect\hyperlink{ref-SjGoBywE}{36}}.
In this context, several unsupervised approaches have been proposed to infer these gene-gene connections by extracting gene modules from co-expression patterns\textsuperscript{\protect\hyperlink{ref-S4e4WaP3}{37}--\protect\hyperlink{ref-VzZoy0BD}{39}}.
Matrix factorization techniques like independent or principal component analysis (ICA/PCA) have shown superior performance in this task\textsuperscript{\protect\hyperlink{ref-1BVbSrr6M}{40}} since they capture local expression effects from a subset of samples and can handle modules overlap effectively.
Therefore, integrating genetic studies with gene modules extracted using unsupervised learning could further improve our understanding of disease origin\textsuperscript{\protect\hyperlink{ref-SjGoBywE}{36}} and progression\textsuperscript{\protect\hyperlink{ref-S56q1qoc}{41}}.

Here we propose PhenoPLIER, an omnigenic approach that provides a gene module perspective to genetic studies.
The flexibility of our method allows integrating different data modalities into the same representation for a joint analysis.
In this work, we show that this module perspective can infer how groups of functionally-related genes influence complex traits, detect shared and distinct transcriptomic properties among traits, and predict how pharmacological perturbations affect genes' activity to exert their effects.
PhenoPLIER maps gene-trait associations and drug-induced transcriptional responses into a common latent representation.
For this, we integrated thousands of gene-trait associations (using TWAS from PhenomeXcan\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42}}) and transcriptional profiles of drugs (from LINCS L1000\textsuperscript{\protect\hyperlink{ref-F7lIlh2N}{43}}) into a low-dimensional space learned from public gene expression data on tens of thousands of RNA-seq samples (recount2\textsuperscript{\protect\hyperlink{ref-6SPTvFXq}{19},\protect\hyperlink{ref-14rnBunuZ}{44}}).
We used a latent representation defined by a matrix factorization approach\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44},\protect\hyperlink{ref-Ki2ij7zE}{45}} that extracts gene modules with certain sparsity constraints and preferences for those that align with prior knowledge (pathways).
When mapping gene-trait associations to this reduced expression space, we observed that diseases were significantly associated with gene modules expressed in relevant cell types: such as hypothyroidism with T cells, corneal endothelial cells with keratometry measurements, hematological assays on specific blood cell types, plasma lipids with adipose tissue, and neuropsychiatric disorders with different brain cell types.
Moreover, since PhenoPLIER can use models derived from large and heterogeneous RNA-seq datasets, we could also identify modules associated with cell types under specific stimuli or disease states.
We observed that significant module-trait associations in PhenomeXcan (our discovery cohort) replicated in the Electronic Medical Records and Genomics (eMERGE) network phase III\textsuperscript{\protect\hyperlink{ref-gZAOkumx}{27},\protect\hyperlink{ref-wfqjCerX}{46}} (our replication cohort).
Furthermore, we performed a CRISPR screen to analyze lipid regulation in HepG2 cells.
We observed more robust trait associations with modules than with individual genes, even when single genes known to be involved in lipid metabolism did not reach genome-wide significance.
Compared to a single-gene approach, our module-based method also better predicted FDA-approved drug-disease links by capturing tissue-specific pathophysiological mechanisms linked with the mechanism of action of drugs (e.g., niacin with cardiovascular traits via a known immune mechanism).
This improved drug-disease prediction suggested that modules may provide a better means to examine drug-disease relationships than individual genes.
Finally, exploring the phenotype-module space revealed stable trait clusters associated with relevant tissues, including a complex branch involving lipids with cardiovascular, autoimmune, and neuropsychiatric disorders.
In summary, instead of considering single genes associated with different complex traits, PhenoPLIER incorporates groups of genes that act together to carry out different functions in specific cell types.
This approach improves robustness in detecting and interpreting genetic associations, and here we show how it can prioritize alternative and potentially more promising candidate targets even when known single gene associations are not detected.
The approach represents a conceptual shift in the interpretation of genetic studies.
It has the potential to extract mechanistic insight from statistical associations to enhance the understanding of complex diseases and their therapeutic modalities.

@@

\begin{figure}
\hypertarget{fig:entire_process}{%
\centering
\includesvg[width=1\textwidth,height=\textheight]{images/entire_process/entire_process.svg}
\caption{\textbf{Schematic of the PhenoPLIER framework.}
\textbf{a)} High-level schematic of PhenoPLIER (a gene module-based method) in the context of TWAS (single-gene) and GWAS (single-variant).
In GWAS, we identify variants associated with traits.
In TWAS, first, we identify variants that are associated with gene expression levels (eQTLs); then, prediction models based on eQTLs are used to impute gene expression, which is used to compute gene-trait associations.
Resources such as LINCS L1000 provide information about how a drug perturbs gene expression; at the bottom-right corner, we show how a drug downregulates two genes (A and C).
In PhenoPLIER, these data types are integrated using groups of genes co-expressed across one or more conditions (such as cell types) that we call gene modules or latent variables/LVs.
\textbf{b)} The integration process in PhenoPLIER uses low-dimensional representations (matrices \(\mathbf{Z}\) and \(\mathbf{B}\)) learned from large gene expression datasets (top).
We used gene-drug information \(\mathbf{L}\) from LINCS L1000 and gene-trait associations \(\mathbf{M}\) from TWAS: PhenomeXcan was used as the discovery cohort, and eMERGE as replication (middle).
PhenoPLIER provides three computational components (bottom):
1) an LV-based regression model that associates an LV \(j\) (\(\mathbf{Z}_j\)) with a trait \(i\) (\(\mathbf{M}_i\)),
2) a clustering framework that learns groups of traits from TWAS associations projected into the LV space (\(\hat{\mathbf{M}}\)),
and 3) an LV-based drug repurposing approach that uses the projection of TWAS (\(\hat{\mathbf{M}}\)) and LINCS L1000 (\(\hat{\mathbf{L}}\)) into the LV space.
\textbf{c)} Genes that are part of LV603, termed as a neutrophil signature\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44}}, were expressed in relevant cell types (top).
LV603 was associated in PhenoPLIER with neutrophil counts and other white blood cells (bottom, showing the top 10 traits for LV603 after projecting gene-trait associations in PhenomeXcan).
eQTLs: expression quantitative trait loci;
MVN: multivariate normal distribution;
PBMC: peripheral blood mononuclear cells;
mDCs: myeloid dendritic cells.}\label{fig:entire_process}
}
\end{figure}

@@

We used TWAS results from PhenomeXcan\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42}} and the eMERGE network\textsuperscript{\protect\hyperlink{ref-gZAOkumx}{27}} as discovery and replication cohorts, respectively (\protect\hyperlink{sec:methods:twas}{Methods}).
PhenomeXcan provides gene-trait associations for 4,091 different diseases and traits from the UK Biobank\textsuperscript{\protect\hyperlink{ref-nmJxPpE5}{47}} and other studies, whereas the analyses on eMERGE were performed across 309 phecodes.
TWAS results were derived using two statistical methods (see \protect\hyperlink{sec:methods:predixcan}{Methods}):
1) Summary-MultiXcan (S-MultiXcan) associations were used for the regression and clustering components,
and 2) Summary-PrediXcan (S-PrediXcan) associations were used for the drug repurposing component.
In addition, we also used colocalization results, which provide a probability of overlap between the GWAS and eQTL signals.
For the drug-repurposing approach, we used transcriptional responses to small molecule perturbations from LINCS L1000\textsuperscript{\protect\hyperlink{ref-F7lIlh2N}{43}} comprising 1,170 compounds.

The latent gene expression representation was obtained from the MultiPLIER models\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44}}, which were derived by applying a matrix factorization method (the pathway-level information extractor or PLIER\textsuperscript{\protect\hyperlink{ref-Ki2ij7zE}{45}}) to recount2\textsuperscript{\protect\hyperlink{ref-6SPTvFXq}{19}} -- a uniformly-curated collection of transcript-level gene expression quantified by RNA-seq in a large, diverse set of samples collected across a range of disease states, cell types differentiation stages, and various stimuli (see \protect\hyperlink{sec:methods:multiplier}{Methods}).
The MultiPLIER models extracted 987 LVs by optimizing data reconstruction but also the alignment of LVs with prior knowledge/pathways.

Each LV or gene module represents a group of weighted genes expressed together in the same tissues and cell types as a functional unit.
Since LVs might represent a functional set of genes regulated by the same transcriptional program\textsuperscript{\protect\hyperlink{ref-1FVd2WW6G}{48},\protect\hyperlink{ref-DN7TyZzb}{49}}, we conjecture that the projection of TWAS and pharmacologic perturbations data into this latent space could provide a better mechanistic understanding.
For this projection of different data modalities into the same space, PhenoPLIER converts gene associations to an LV score: all genes' standardized effect sizes for a trait (from TWAS) or differential expression values for a drug (from pharmacologic perturbation data) are multiplied by the LV genes' weights and summed, producing a single value.
Instead of looking at individual genes, this process links different traits and drugs to functionally-related groups of genes or LVs.
PhenoPLIER uses LVs annotations about the specific conditions where the group of genes is expressed, such as cell types and tissues, even at specific developmental stages, disease stages or under distinct stimuli.
Although this is not strictly necessary for PhenoPLIER to work, these annotations can dramatically improve the interpretability of results.
MultiPLIER's models provide this information by linking LVs to samples, which may be annotated for experimental conditions (represented by matrix \(\mathbf{B}\) at the top of Figure \ref{fig:entire_process}b) in which genes in an LV are expressed.
An example of this is shown in Figure \ref{fig:entire_process}c.
In the original MultiPLIER study, the authors reported that one of the latent variables, identified as LV603, was associated with a known neutrophil pathway and highly correlated with neutrophil count estimates from whole blood RNA-seq profiles\textsuperscript{\protect\hyperlink{ref-18TSqd1tG}{50}}.
We analyzed LV603 using PhenoPLIER and found that
1) neutrophil counts and other white blood cell traits were ranked among the top 10 traits out of 4,091 (Figure \ref{fig:entire_process}c, bottom), and basophils count and percentage were significantly associated with this LV when using our regression method (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv603}),
and 2) LV603's genes were expressed in highly relevant cell types (Figure \ref{fig:entire_process}c, top).
These initial results suggested that groups of functionally related and co-expressed genes tend to correspond to groups of trait-associated genes, and the approach can link transcriptional mechanisms from large and diverse dataset collections to complex traits.
